---
title: "CSIS3360-Project"
author: "300360903 - 300353099"
date: "2023-08-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



#STEP 1 - Load all the necessary libraries
```{r}

library(stringr)
library(data.table)
library(MASS)

```


#STEP 2 -Read and Process the data
```{r}

salaries_df <- read.csv("salaries.csv", header = TRUE, sep = ",")

head(salaries_df)



```

change 'Year' column to chr data type
'ConvertedComp' column to int data type
```{r}

salaries_df$Year <- as.character(salaries_df$Year)

salaries_df$ConvertedComp <- as.integer(salaries_df$ConvertedComp)

salaries_df_copy <- salaries_df


```


================================================================================
================================================================================
================================================================================
================================================================================
================================================================================


                            
                                REGRESSION


#PART 1 - Regression (salary prediction)


Job satisfaction cannot predict salary...
Salary determines how satisfied a person is, in their job...

```{r}

salaries_df <- subset(salaries_df, select = -c(JobSat))

head(salaries_df)

```

##Change all column names to lower case
```{r}

colnames(salaries_df) <- tolower(colnames(salaries_df))

head(salaries_df)

```

================================================================================

##Process Categorical columns

--------------------------------------------------------------------------------
Change 'hobbyist' column to have yes or no values...
```{r}

hobby <- function(x){
  if (str_sub(x, end = 3) == "Yes")
  {
    return ("yes")
  }
  else if (str_sub(x, end = 2) == "No")
  {
    return ("no")
  }
}

salaries_df$hobbyist <- sapply(salaries_df$hobbyist,hobby)

```


```{r}

table(salaries_df$hobbyist)

```

--------------------------------------------------------------------------------
Change 'country' column to have north america or other as values

..This is because the salaries are converted to US dollars and are
higher in the north american continent, compared to other places..
[United States, Canada] -- North America
[Every other value] -- Other
```{r}

country <- function(x){
  if (x == "United States" || x == "Canada")
  {
    return ("North_America")
  }
  else
  {
    return ("Other")
  }
}

salaries_df$country <- sapply(salaries_df$country, country)

```


```{r}

table(salaries_df$country)

```

--------------------------------------------------------------------------------
Change 'edlevel' column to have
[Bachelor's degree] -- Bachelors
[Master's degree] -- Masters
[Doctoral degree] -- PhD
[Other degree values] -- Other
[No formal education] -- No_education
```{r}

edlevel <- function(x) {
  if (x %in% c("Associate degree", "Professional degree")) 
  {
    return ("Certificate_studies")
  } 
  else if (x == "Bachelor's degree") 
  {
    return ("Bachelors")
  } 
  else if (x == "Master's degree") 
  {
    return ("Masters")
  } 
  else if (x == "Doctoral degree") 
  {
    return ("PhD")
  } 
  else 
  {
    return ("No_professional_education")
  }
}

salaries_df$edlevel <- sapply(salaries_df$edlevel, edlevel)

```


```{r}

table(salaries_df$edlevel)

```

--------------------------------------------------------------------------------
Change 'employment' column to have
[Employed full-time] -- full_time
[Employed part-time] -- part_time
[every other value] -- self_employed
```{r}

employment <- function(x) {
  if (x == "Employed full-time") 
  {
    return ("full_time")
  } 
  else if (x == "Employed part-time") 
  {
    return ("part_time")
  } 
  else 
  {
    return ("self_employed")
  }
}

salaries_df$employment <- sapply(salaries_df$employment, employment)

```


```{r}

table(salaries_df$employment)

```

--------------------------------------------------------------------------------
change 'orgsize' column to have
[0-100 employees] -- small
[100-1000 employees] -- medium
[>1000 employees] -- large
[every other value] -- unknown
```{r}

orgsize <- function(x) {
  if (x %in% c("20 to 99 employees", "2 to 9 employees", "10 to 19 employees", "Just me - I am a freelancer, sole proprietor, etc.")) 
  {
    return ("small")
  } 
  else if (x %in% c("100 to 499 employees", "500 to 999 employees")) 
  {
    return ("medium")
  } 
  else if ( x %in% c("I don't know", "I prefer not to answer", "0"))
  {
    return ("unknown")
  }
  else 
  {
    return ("large")
  }
}

salaries_df$orgsize <- sapply(salaries_df$orgsize, orgsize)

```

```{r}

table(salaries_df$orgsize)

```

--------------------------------------------------------------------------------

change 'undergradmajor' column to have

[Computer science,Information systems,
Computer engineering or electrical/electronics engineering,
Computer programming or Web development] -- Relevant_to_job

[100-1000 employees] -- Not_relevant
```{r}

undergrad <- function(x) {
  if (x %in% c("Computer science", "Information systems", "Computer engineering or electrical/electronics engineering", "Computer programming or Web development")) 
  {
    return ("Relevant_to_Job")
  } 
  else
  {
    return ("Not_Relevant")
  }
}

salaries_df$undergradmajor <- sapply(salaries_df$undergradmajor, undergrad)

```

```{r}

table(salaries_df$undergradmajor)

```

--------------------------------------------------------------------------------

change 'yearscodepro' column to have

[<=5] -- beginner
[5-10] -- associate
[10-15] -- senior
[>15] -- expert
```{r}

experience <- function(x) {
  if (x <= 5) {
    return ("beginner")
  } 
  else if (x > 5 && x <= 10) 
  {
    return ("associate")
  } 
  else if (x >= 10 && x < 15) 
  {
    return ("senior")
  } 
  else 
  {
    return("expert")
  }
}

salaries_df$yearscodepro <- sapply(salaries_df$yearscodepro, experience)

```

```{r}

table(salaries_df$yearscodepro)

```


```{r}

head(salaries_df)

```

##Separate categorical columns and create dummies
```{r}

categorical_cols <- subset(salaries_df, select = -c(3,10,11,12,13))

cat_cols_dummy <- model.matrix(~., data = categorical_cols)

cat_cols_dummy <- subset(cat_cols_dummy, select = -c(1))

head(cat_cols_dummy)

```

##Combine dummies with df
```{r}

final_regression_df <- cbind(subset(salaries_df, select = c(3,10,11,12,13)),cat_cols_dummy)

head(final_regression_df)

```

#Explanatory Model - Regression
```{r}

options(scipen = 99)
#To suppress scientific notations...

linear_reg <- lm(convertedcomp ~. , data = final_regression_df)

summary(linear_reg)

```

================================================================================

##Interpretation:
Coefficients for each variables shows how they are associated with
'convertedcomp' as explained below:

--------------------------------------------------------------------------------

SIGNIFICANCE : Looking at the p-values, all the factors are shown to be 
statistically significant in determining the salary of an individual. 
But, out of the given factors, an individual not having any formal education 
does not have as much significance as the other factors...So, it is unlikely 
to have occurred by random chance.

--------------------------------------------------------------------------------

SIGN :
data_scientist - Positively Associated
database.administrator - Negatively Associated
data_analyst - Negatively Associated
data_engineer - Positively Associated
year2018 - Positively Associated
year2019 - Positively Associated
year2020 - Positively Associated
hobbyistyes - Positively Associated
countryOther - Negatively Associated
edlevelCertificate_studies - Negatively Associated
edlevelMasters - Positively Associated
edlevelNo_professional_education - 
edlevelPhD - Positively Associated
employmentpart_time - Negatively Associated
employmentself_employed - Positively Associated
orgsizemedium - Negatively Associated
orgsizesmall - Negatively Associated
orgsizeunknown - Negatively Associated
undergradmajorRelevant_to_Job - Negatively Associated
yearscodeprobeginner - Negatively Associated
yearscodeproexpert - Positively Associated
yearscodeprosenior - Positively Associated



--------------------------------------------------------------------------------

SIZE :

- being a data_scientist increases salary by $3736.10 from the average...
- being a database.administrator decreases salary by $5115.30 from the average...
- being a data_analyst decreases salary by $4890.70 from the average...
- being a data_engineer increases salary by $8473.50 from the average...
- compared to 2017 the average salary in the year 2018 was higher by $8732.50...
- compared to 2017 the average salary in the year 2019 was higher by $7539.80...
- compared to 2017 the average salary in the year 2020 was higher by $6926.70...
- if the person codes outside of work as hobby, he tends to have a salary of
  $2909.80 more than someone who does not...
- A person employed outside the United States or Canada on an average tend to
  have a salary of $48,913.30 less... This reflects the highly competitive 
  salaries in the North American continent...
- A person who completed only certifications tends to have a salary of $8238 less
  than somebody who completed their Bachelors Degree....
- A person who completed a Masters Degree tends to have a salary of $6270.50 more
  than somebody who completed their Bachelors Degree....
- A person who completed no formal education tends to have a salary of $2181.60 less
  than somebody who completed their Bachelors Degree.... 
- A person who completed a Doctoral Degree tends to have a salary of $16,042.30 more
  than somebody who completed their Bachelors Degree....
- A person who is employed Part-Time tends to have a salary of $16,056 less
  than somebody who is employed Full-Time....
- A person who is self-employed tends to have a salary of $8570.40 more
  than somebody who is employed Full-Time....
- A person who is employed in a medium size organization tends to have a salary 
  of $5298.10 less than somebody who is employed in a large organization....
- A person who is employed in a small size organization tends to have a salary 
  of $8359.30 less than somebody who is employed in a large organization....
- A person who does not know the size of his/her organization tends to have a salary 
  of $9695.80 less than somebody who is employed in a large organization....
- A person who’s undergrad major was relevant to the job, tends to earn
  $3083 less on average compared to someone with an irrelevant degree...
  This does not make sense in the real world but might be because a person
  with an irrelevant degree is hired into a technical job because of his/her
  domain expertise, which is valued more than technical knowledge in the 
  Data field.
- A beginner with less than 5 years of experience earns $17,824.10 less than 
  an associate with 5-10 years of experience...
- An expert with more than 15 years of experience earns $25,077.40 more than 
  an associate with 5-10 years of experience...
- A senior with 10-15 years of experience earns $11,901.10 more than an 
  associate with 5-10 years of experience...



Except for the coefficient for the column 'undergradmajorRelevant_to_Job' being 
negatively associated, every other factors' coefficient makes sense in the real 
world. But, the accuracy of the model tends to be low, because even when all the 
factors are the same, the salary tends to vary a lot in the real world. This makes
it difficult for prediction.


--------------------------------------------------------------------------------
--------------------------------------------------------------------------------
--------------------------------------------------------------------------------
--------------------------------------------------------------------------------

#Perform prediction using Linear Regression Model

Since we are Data Analysts, we predict the salary for a beginner
data analyst in North America in the year 2019 as it was the year before covid

```{r}

data_analyst_beginner <- data.frame(data_scientist = 0,
                          database.administrator = 0,
                          data_analyst = 1,
                          data_engineer = 0,
                          year2018 = 0,
                          year2019 = 1,
                          year2020 = 0,
                          hobbyistyes = 1,
                          countryOther = 0,
                          edlevelCertificate_studies = 0,
                          edlevelMasters = 1,
                          edlevelNo_professional_education = 0,
                          edlevelPhD = 0,
                          employmentpart_time = 0,
                          employmentself_employed = 0,
                          orgsizemedium = 0,
                          orgsizesmall = 0,
                          orgsizeunknown = 0,
                          undergradmajorRelevant_to_Job = 0,
                          yearscodeprobeginner = 1,
                          yearscodeproexpert = 0,
                          yearscodeprosenior = 0
                          )


predicted_salary <- predict(linear_reg, data_analyst_beginner, interval = "prediction",
                        level = 0.05)

predicted_salary

```

A beginner Data Analyst with the above given specifications is predicted to 
have a salary of $89,841.33.

--------------------------------------------------------------------------------

```{r}

data_analyst_beginner_selfEmp <- data.frame(data_scientist = 0,
                          database.administrator = 0,
                          data_analyst = 1,
                          data_engineer = 0,
                          year2018 = 0,
                          year2019 = 1,
                          year2020 = 0,
                          hobbyistyes = 1,
                          countryOther = 0,
                          edlevelCertificate_studies = 0,
                          edlevelMasters = 1,
                          edlevelNo_professional_education = 0,
                          edlevelPhD = 0,
                          employmentpart_time = 0,
                          employmentself_employed = 1,
                          orgsizemedium = 0,
                          orgsizesmall = 0,
                          orgsizeunknown = 0,
                          undergradmajorRelevant_to_Job = 0,
                          yearscodeprobeginner = 1,
                          yearscodeproexpert = 0,
                          yearscodeprosenior = 0
                          )


predicted_salary <- predict(linear_reg, data_analyst_beginner_selfEmp, interval = "prediction",
                        level = 0.05)

predicted_salary

```

If he/she lands a contract job, or works as a Data Analytics consultant,
and has expert knowledge without the experience, the salary is predicted 
to be $98,411.69.


--------------------------------------------------------------------------------

```{r}

data_analyst_associate <- data.frame(data_scientist = 0,
                          database.administrator = 0,
                          data_analyst = 1,
                          data_engineer = 0,
                          year2018 = 0,
                          year2019 = 1,
                          year2020 = 0,
                          hobbyistyes = 1,
                          countryOther = 0,
                          edlevelCertificate_studies = 0,
                          edlevelMasters = 1,
                          edlevelNo_professional_education = 0,
                          edlevelPhD = 0,
                          employmentpart_time = 0,
                          employmentself_employed = 0,
                          orgsizemedium = 0,
                          orgsizesmall = 0,
                          orgsizeunknown = 0,
                          undergradmajorRelevant_to_Job = 0,
                          yearscodeprobeginner = 0,
                          yearscodeproexpert = 0,
                          yearscodeprosenior = 0
                          )


predicted_salary <- predict(linear_reg, data_analyst_associate, interval = "prediction",
                        level = 0.05)

predicted_salary

```

After around 5 years of experience his salary is predicted to 
be $107,665,50.

--------------------------------------------------------------------------------

```{r}

data_analyst_beginner_outOfCountry <- data.frame(data_scientist = 0,
                          database.administrator = 0,
                          data_analyst = 1,
                          data_engineer = 0,
                          year2018 = 0,
                          year2019 = 1,
                          year2020 = 0,
                          hobbyistyes = 1,
                          countryOther = 1,
                          edlevelCertificate_studies = 0,
                          edlevelMasters = 1,
                          edlevelNo_professional_education = 0,
                          edlevelPhD = 0,
                          employmentpart_time = 0,
                          employmentself_employed = 0,
                          orgsizemedium = 0,
                          orgsizesmall = 0,
                          orgsizeunknown = 0,
                          undergradmajorRelevant_to_Job = 0,
                          yearscodeprobeginner = 1,
                          yearscodeproexpert = 0,
                          yearscodeprosenior = 0
                          )


predicted_salary <- predict(linear_reg, data_analyst_beginner_outOfCountry, interval = "prediction",
                        level = 0.05)

predicted_salary

```

If the person looks for a job outside the north american continent
as a beginner, he/she is expected to get a salary of $40,928.04.


================================================================================
================================================================================
================================================================================
================================================================================
================================================================================


            
                            CLASSIFICATION
              

#PART 2 - Classification (based on job satisfaction)

```{r}

job_sat_df <- read.csv("salaries.csv", header = TRUE, sep = ",")

final_classification_df <- cbind(final_regression_df, job_sat_df[,c(7)])


final_classification_df <- final_classification_df %>% setnames(old = "job_sat_df[, c(7)]", 
                                                                new = "job_satisfaction")

head(final_classification_df)


 
```

--------------------------------------------------------------------------------
Change 'job_satisfaction' column to have
[<=6] -- less_satisfied
[>6] -- satisfied

```{r}

job_sat <- function(x){
  if (x <= 6) {
    return ("less_satisfied")
  } 
  else 
  {
    return("satisfied")
  }
}

final_classification_df$job_satisfaction <- sapply(final_classification_df$job_satisfaction, job_sat)

```


```{r}

table(final_classification_df$job_satisfaction)

```

##Create dummies for job satisfaction column
```{r}

job_sat_col <- subset(final_classification_df, select = c("job_satisfaction"))

job_sat_dummy <- model.matrix(~., data = job_sat_col)

job_sat_dummy <- subset(job_sat_dummy, select = -c(1))

head(job_sat_dummy)

```

##Combine it with the dataframe
```{r}

final_classification_df <- cbind(subset(final_classification_df, select = -c(24)), job_sat_dummy)

head(final_classification_df)

```

##Logistic regression Model
```{r}

logistic_reg = glm(job_satisfactionsatisfied ~., data = final_classification_df,
                   family = binomial(link = "logit"))


summary(logistic_reg)

```

##inrepretation:
Deviance Residuals represent that it should be normally distributed around 
zero by the range of -2.0129 to 1.8485.

Positive coefficients indicate an increase in the job satisfaction.

The difference of Null Deviance and Residual Deviance show how well the model 
is fitted by smaller value after fitting the model.



###odds ratio
```{r}

cbind(OddsRatio = exp(coef(logistic_reg)), exp(confint(logistic_reg)))

```

##Interpretation:
These odds ratio represent insights into how different variables are associated 
with job satisfaction. For example, Being a data scientist increases the 
odds of the person being satisfied with their job by 16.35%, or 
having completed a PhD increases the odds of the person being satisfied with 
their job by 33.59% because the salary tends to be way higher than their peers..


#refining logistic regression model

```{r}

stepwise_model <- stepAIC(logistic_reg, direction = "backward",
                          trace = FALSE)

summary(stepwise_model)

```
##Interpretation:
The range of residuals is from -1.9983 to 1.8525. It shows better normally
distribution around zero that before.

A one-unit increase in salary (convertedcomp) is associated with an increase of 
approximately 0.000005933 in the log-odds of job satisfaction.
Belonging to "Other" countries is associated with a decrease of approximately
0.119 in the log-odds of job satisfaction.
Having a Master's or PhD degree is associated with changes in the log-odds of 
job satisfaction, but only edlevelPhD is statistically significant.

Smaller values of Residual deviance in comparison of Null deviance suggests a 
better fit.  and indicating an improvement in model fit.

AIC value is decreased compared to the previous model, representing improved 
model performance.


###odds ratio
```{r}

cbind(OddsRatio = exp(coef(stepwise_model)), exp(confint(stepwise_model)))

```
##Interpretation:
These odds ratio represent insights into how different variables are associated 
with job satisfaction. For example, Being a data scientist increases the 
odds of the person being satisfied with their job by 15.73%, or 
having completed a PhD increases the odds of the person being satisfied with 
their job by 32.44% because the salary tends to be way higher than their peers...
These odds are from the improved model...

Looking at the odds ratios over the years the job satisfaction tends to reduce...
This might be because of increasing competitveness in the Data field..
(i.e) The odds of a person being satisfied with his/her job in the year 2020,
drastically dropped by around 75% compared to 2017...



======================================****======================================
